A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods

### 2024-01-04 Draft/submission1/Draft_MEM.tex
Adding files as of submission 1.

\documentclass{article}
\usepackage[left=2.5cm,right=2.5cm, top = 2.5cm, bottom = 3cm]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{amsmath}

@@

% \title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
\title{A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods}
\author{Johannes Bracher$^{1, 2}$, Jonas M. Littek$^1$}
\date{9 June 2023\\ \medskip \small
$^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
$^2$Heidelberg Institute for Theoretical Studies (HITS), Computational Statistics Group}

@@

\begin{center}
\textbf{This manuscript is is a preprint and has not been subject to peer review.}
\end{center}

\abstract{
\begin{quote}
\footnotesize Intensity thresholds serve to make influenza activity comparable across countries and surveillance systems. The moving epidemic method (MEM) and the WHO method are widely used to classify season peaks as low, medium, high or very high. In both methods, thresholds correspond to quantiles of a normal distribution fitted to a set of potentially transformed historical observations. However, there are differences in how this set is constructed. We assess the impact of these choices analytically and in simulation studies. We find that under the default settings of both the MEM and WHO method, on average more season peaks than intended are classified as high or very high. Combining characteristics of both, better-calibrated thresholds can be achieved. Even these, however, must be expected to have rather modest sensitivity and positive predictive values.
\end{quote}
}

@@

{\begin{quote}
\footnotesize Intensity thresholds are a common tool to make influenza activity comparable across different countries and surveillance systems. The moving epidemic method (MEM) and the WHO method are widely used to this end and enable classification of season peak intensity as low, medium, high or very high. The two approaches are similar in that they base thresholds on quantiles of a normal distribution fitted to a reference set of historical observations. However, three methodological differences exist. Firstly, in the MEM the normal distribution is fitted to log-transformed incidence data, while the WHO method by default operates on the original scale. Secondly, the MEM uses more than one observation from each past season, fixing the total number to include into the reference set. The WHO method uses only the highest value from each season. Lastly, in the WHO method, but not the MEM, historical data are by default  smoothed prior to the computation of thresholds. We assess the impact of these choices on thresholds both analytically and in a simulation study. The latter is based on re-sampling of influenza-like illness (ILI) data from France and the US, thus reflecting temperate climate settings. We find that when the normal distribution is fitted to untransformed observations, a rather large proportion of new season peaks are classified as high or very high intensity. This can be mitigated by a logarithmic transformation. When fixing the total number of past observations included in the reference set as in the MEM, thresholds increase in expectation the more seasons are available. When only few are available, there is a high chance of classifying new season peaks as high or very high intensity. Smoothing incidence time series prior to computing thresholds results in somewhat less variable estimators, but also a lowering of thresholds. If these are applied to unsmoothed new season peaks, there will again be a large proportion classified as high or very high intensity. If they are applied to smoothed new data, this problem is avoided. In terms of sensitivity and positive predictive values of thresholds, we find that these cannot be expected to exceed rather modest levels if the number of available historical seasons is low; thresholds for very high intensity are particularly affected by this problem. Our practical recommendation is to include one observation per season into the reference set and employ a log transformation in the computation of thresholds. Smoothing can be applied to somewhat reduce the variability of thresholds. However, threholds then need to be applied to smoothed rather than raw new peak values, which slightly modifies their interpretation.
\end{quote}}

@@

\includegraphics[width=0.8\textwidth]{figure/heat_chart.png}
\end{center}
\caption{Influenza intensity map (top, week 01/2023) and heat chart (bottom, weeks 01--07/2023) as published by \textit{Flu News Europe} (\url{https://flunewseurope.org/SeasonOverview}). Copyright: World Health Organization, 2023, and European Centre for Disease Prevention and Control, 2023.
Reproduction is authorised, provided the source is acknowledged}
\label{fig:maps}
\end{figure}

@@

\noindent We note that while the overarching goal of the \textit{WHO PISA} guidelines is to enhance preparedness for \textit{pandemic influenza}, the intended statistical properties of the MEM and WHO methods refer to \textit{seasonal influenza waves}. The rationale is that a good understanding of the range of peak intensities of seasonal influenza is the basis for a meaningful assessment of pandemic influenza outbreaks. Our reasoning in the following therefore focuses on the setting of subsequent waves of seasonal influenza. Note that we only address the question of peak intensity thresholds, but not season onset / baseline thresholds \citep{Vega2013}. Also, we do not address the question of determining peak intensity in real time, i.e., before the season peak has been reached or before it is clear whether a peak has been reached. Rather than simple thresholds, this would require control charts (see e.g., \citealt{Liu2019}), which we consider outside the scope of the present article.

@@

\begin{enumerate}
\item \textbf{Smoothing of historical data} (optional): apply an $l$-week moving average to all historical seasons. If data are smoothed, we denote by $x^\text{raw}_{j, k}$ the raw observation from season $j = 1, \dots, m$, week $k = 1, \dots, 52$, and by
$$
x^\text{smo}_{j, k} = \frac{1}{l} \sum_{d = 0}^{l - 1} x^\text{raw}_{j, k - d}, \ \ k = l, \dots, 52
$$ the smoothed version. In the remainder of this description we denote whichever of $x^\text{raw}_{j, k}$ and $x^\text{smo}_{j, k}$ is used to compute thresholds by $x_{j, k}$.
\item \textbf{Sorting:} Within each historical season $j = 1, \dots, m$ order all observations $x_{j, k}$ in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(i)}_j$. More generally we will in the following use the notation $^{(\cdot)}$ to denote an ordering of values, e.g., denoting by $x^{\text{raw}, (i)}_j$ the $i$-th largest unsmoothed incidence value from season $j$.
\item \textbf{Selection of reference set:} Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_j^{(i)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item \textbf{Data transformation:} Apply a monotonically increasing transformation $y_j^{(i)} = f(x_j^{(i)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
\item \textbf{Fitting a normal distribution:} Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates of its mean and standard deviation,
\begin{align}
\begin{split}
\bar{y} & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n y_j^{(i)}}{nm}\\
\sd & = \sqrt{ \frac{\sum_{j = 1}^m\sum_{i = 1}^n (y_j^{(i)}  - \bar{y})}{nm - 1}}.
\end{split}\label{eq:moments}
\end{align}
\item \textbf{Computation of thresholds:} Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\bar{y}, \sd^2)$, i.e.\ compute
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd, \label{eq:def_q}
\end{equation}
where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution. The $\hat{q}_{Y, \alpha}$ can be interpreted as estimates of quantiles $q_{Y, \alpha}$ of an underlying theoretical distribution. A common choice for $\alpha$ is
\begin{itemize}
\item the 40th percentile $\hat{q}_{Y, 0.4} = \bar{y} - 0.25 \sd$ as the threshold for medium intensity;
\item the 90th percentile $\hat{q}_{Y, 0.9} = \bar{y} + 1.28 \sd$ as the threshold for high intensity;
\item the 97.5th percentile $\hat{q}_{Y, 0.975} = \bar{y} + 1.96\sd$ as the threshold for very high intensity.
\end{itemize}
\item \textbf{Transformation of thresholds to original scale:} Obtain corresponding thresholds on the original scale by applying the inverse transformation, i.e. setting % $\hat{q}_{X, 0.4} = f^{-1}(\hat{q}_{Y, 0.4})$, $\hat{q}_{X, 0.9} = f^{-1}(\hat{q}_{Y, 0.9})$, $\hat{q}_{X, 0.975} = f^{-1}(\hat{q}_{Y, 0.975})$.
$$
\hat{q}_{X, \alpha} = f^{-1}(\hat{q}_{Y, \alpha})
$$
for $\alpha = 0.4, 0.9, 0.975$.
\item \textbf{Application of thresholds:} The obtained thresholds are applied to classify the season peak value of the $(m + 1)-th$ season. Depending on the exact specification, thresholds can be applied either to the raw peak value $x_{m + 1}^{\text{raw}, (1)}$ or the smoothed version $x_{m + 1}^{\text{smo}, (1)}$.
\end{enumerate}

@@

The implementation of the WHO method in the publicly available \textit{WHO Average Curves} application \citep{WHO2023} likewise offers a lot of flexibility; our description is based on the default settings as of May 2023, as well as the description in \cite{WHO2014}. Smoothing of data prior to the computation of thresholds is recommended, with a default of $l = 3$ (adapted from a recommendation of $l = 4$ in \citealt[p68]{WHO2014}). Subsequently, $n = 1$ observation per season is used and by default no transformation is applied to the reference set. If peak incidences vary strongly across seasons, a log transformation is recommended. At least three historical seasons are required to compute thresholds, but it is noted that the ``accuracy of these thresholds should be expected to increase with the number of seasons of good quality data available'' \cite[p22]{WHO2023}. % \citep{WHO2017}.
Thresholds are by default applied to unsmoothed new season peaks.

@@

To improve our understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search of articles published in English language and citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}). The results are summarized in Table \ref{tab:literature}. As can be seen from the large number of entries from the years 2019 and 2020, the MEM has quickly become a standard approach in the determination of intensity thresholds for influenza and other respiratory diseases. The articles come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies. In most analyses, threshold levels at the 40th, 90th and 97.5th percentile as in Section \ref{sec:definitions} are used. Variability with respect to the number $m$ of historical seasons included is considerable, with a range from 3 to 16 seasons. Consequently, the number $n$ of observations included per season ranges from two to ten (none of the above papers indicated a modification of the default setting $n = 30/m$ of the moving epidemic method). We only found three published applications of the WHO method, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.

@@

\caption{\footnotesize Published applications of the MEM and WHO methods to estimate intensity thresholds. The number of seasons used to compute thresholds is denoted by $m$, the number of observations used per season by $n$. ``Percentiles'' indicates which percentiles were used for the medium, high and very high thresholds, with ``?'' indicating no information was found. Abbreviations: ARI/SARI = (severe) acute respiratory infection; ILI = influenza-like illness; RSV = respiratory syncytial virus.}

@@

@@

Maritius & ARI & 2012--2018 & varies$^1$ & varies & 40, 90, 97.5 & Teeluck et al (\citeyear{Teeluck2021}) \\

@@

\midrule \\

@@

\bottomrule\\
\multicolumn{7}{c}{$^1$ Teeluck at al (\citeyear{Teeluck2021}) use various methods to split up multi-peak seasons, leading to different effective numbers of waves used.}\\ 

@@

We first consider the impact of the number $n$ of observations used per historical season. We assume that no smoothing is applied in Step 1 of the algorithm described in Section \ref{sec:definitions}. Denote by $\mathbf{Y}_j$ the random vector of the $n$ largest transformed incidence values from season $j$ in decreasing order, i.e.

@@

We assume that $\mathbf{Y}_1, \dots, \mathbf{Y}_m$ are identically and independently distributed, denoting their theoretical mean and covariance matrix of by

@@

To make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. Construction of thresholds is based on the reference set $\mathcal{Y}$, which pools the elements of the vectors $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_m$. It can be shown that the expectations of the empirical mean $\bar{Y}$ and variance $S^2$ of the observations in the reference set (see in equation \eqref{eq:moments}) are given by

@@

where the second equality holds only if the transformed peak values $Y_{j}^{(1)}$ indeed come from a normal distribution.  This is a desirable property as our estimator $\hat{q}_{Y, \alpha}$ of $q_{Y, \alpha}$ is approximately unbiased. In expectation, thresholds on the transformed scale are thus such that the nominal threshold exceedance probabilities (60\%/10\%/2.5\%) are achieved. If a different value $n > 1$ is chosen, this will generally no longer be the case. Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $\hat{q}_{Y, \alpha}$ can then be expected to differ from $q_{Y, \alpha}$. By the definition of $\mu_i$ as the expectation of the $i$-th largest observation in a given season, $\mathbb{E}(\bar{Y})$ decreases in $n$. While the expected thresholds also depend on $\mathbf{\Sigma}$, this downward bias will usually translate to the $\hat{q}_{Y, \alpha}$. As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity. Intuitively speaking, extending the reference set by observations which are close to peaks, but not actually peaks, dilutes the reference set and pulls thresholds downward. When choosing $n = 30/m$ as suggested for the MEM, thresholds will tend to increase the more years of data are used to compute thresholds, and the probability of exceeding thresholds will decrease (approaching the nominal values from above).

@@

\subsection{Smoothing of time series prior to computing of thresholds}
\label{subsec:smoothing}

>>>>

\documentclass{article}
\usepackage[left=2.5cm,right=2.5cm, top = 2.5cm, bottom = 2.8cm]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{amsmath}

@@

% \title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
\title{A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods}
\author{Johannes Bracher$^{1, 2}$, Jonas M. Littek$^1$}
\date{9 June 2023\\ \medskip \small
$^1$Karlsruhe Institute of Technology (KIT), Chair of Statistical Methods and Econometrics\\
$^2$Heidelberg Institute for Theoretical Studies (HITS), Computational Statistics Group\\ \medskip
Email address for correspondence:\\ \url{johannes.bracher@kit.edu}\\ % \medskip
% Postal address: \\ Karlsruhe Institute of Technology, Institute of Economics,\\ Bl\"ucherstrasse 17, 76185 Karlsruhe, Germany
}

@@

% \begin{center}
% \textbf{This manuscript is is a preprint and has not been subject to peer review.}
% \end{center}

% \begin{center}
% \textbf{Running title:} Assessment of influenza intensity thresholds
% \end{center}

\abstract{
\begin{quote}
\footnotesize Intensity thresholds serve to make influenza activity comparable across countries and surveillance systems. The moving epidemic method (MEM) and the WHO method are widely used to classify season peaks as low, medium, high or very high. In both methods, thresholds correspond to quantiles of a normal distribution, fitted to a set of potentially transformed historical observations. However, several differences between the two procedures exist, which we study analytically and using re-sampling of real-world data. We find that under the default settings, both the MEM and WHO method on average classify more season peaks than intended as high or very high. Combining characteristics of both, better-calibrated thresholds can be achieved. Even these, however, have modest sensitivity and positive predictive values. This concerns especially thresholds for the very high intensity level, which should thus be interpreted with care.
\end{quote}
}

@@

{\begin{quote}
\footnotesize Intensity thresholds are a common tool to make influenza activity comparable across different countries and surveillance systems. The moving epidemic method (MEM) and the WHO method are widely used to this end and enable classification of season peak intensity as low, medium, high or very high. The two approaches are similar in that they base thresholds on quantiles of a normal distribution fitted to a reference set of historical observations. However, three methodological differences exist. Firstly, in the MEM the normal distribution is fitted to log-transformed incidence data, while the WHO method by default operates on the original scale. Secondly, the MEM uses more than one observation from each past season, fixing the total number to include into the reference set. The WHO method uses only the highest value from each season. Lastly, in the WHO method, but not the MEM, historical data are by default  smoothed prior to the computation of thresholds. We assess the impact of these choices on thresholds both analytically and in a simulation study. The latter is based on re-sampling of influenza-like illness (ILI) data from France and the US, thus reflecting temperate climate settings. We find that when the normal distribution is fitted to untransformed observations, a rather large proportion of new season peaks are classified as high or very high intensity. This can be mitigated by a logarithmic transformation. When fixing the total number of past observations included in the reference set as in the MEM, thresholds increase in expectation the more seasons are available. When only few are available, there is a high chance of classifying new season peaks as high or very high intensity. Smoothing incidence time series prior to computing thresholds results in somewhat less variable estimators, but also a lowering of thresholds. If these are applied to unsmoothed new season peaks, there will again be a large proportion classified as high or very high intensity. If they are applied to smoothed new data, this problem is avoided. In terms of sensitivity and positive predictive values of thresholds, we find that these cannot be expected to exceed rather modest levels if the number of available historical seasons is low; thresholds for very high intensity are particularly affected by this problem. Our practical recommendation is to include one observation per season into the reference set and employ a log transformation in the computation of thresholds. Smoothing can be applied to somewhat reduce the variability of thresholds. However, threholds then should be applied to smoothed rather than raw new peak values, which slightly modifies their interpretation.
\end{quote}}

@@

\includegraphics[width=0.8\textwidth]{figure/heat_chart.png}
\end{center}
\caption{Influenza intensity map (top, week 01/2023) and heat chart (bottom, weeks 01--07/2023) as published by \textit{Flu News Europe} (\url{https://flunewseurope.org/SeasonOverview}). Copyright: World Health Organization, 2023, and European Centre for Disease Prevention and Control, 2023.
Reproduction is authorised, provided the source is acknowledged.}
\label{fig:maps}
\end{figure}

@@

\noindent We note that while the overarching goal of the \textit{WHO PISA} guidelines is to enhance preparedness for \textit{pandemic influenza}, the intended statistical properties of the MEM and WHO methods refer to \textit{seasonal influenza waves}. The rationale is that a good understanding of the range of peak intensities of seasonal influenza is the basis for a meaningful assessment of pandemic influenza outbreaks. Our reasoning in the following therefore focuses on the setting of subsequent waves of seasonal influenza. Note that we only address the question of peak intensity thresholds, but not season onset / baseline thresholds \citep{Vega2013}; some analyses on the behaviour of these thresholds can be found in \cite{Pang2023}. Also, we do not address the question of determining peak intensity in real time, i.e., before the season peak has been reached or before it is clear whether a peak has been reached. Rather than simple thresholds, this would require control charts (see e.g., \citealt{Liu2019}), which we consider outside the scope of the present article.

@@

\begin{enumerate}
\item[(a)] \textbf{Smoothing of historical data} (optional): apply an $l$-week moving average to all historical seasons. If data are smoothed, we denote by $x^\text{raw}_{j, k}$ the raw observation from season $j = 1, \dots, m$, week $k = 1, \dots, 52$, and by
$$
x^\text{smo}_{j, k} = \frac{1}{l} \sum_{d = 0}^{l - 1} x^\text{raw}_{j, k - d}, \ \ k = l, \dots, 52
$$ the smoothed version. In the remainder of this description we denote whichever of $x^\text{raw}_{j, k}$ and $x^\text{smo}_{j, k}$ is used to compute thresholds by $x_{j, k}$.
\item[(b)] \textbf{Sorting:} Within each historical season $j = 1, \dots, m$ order all observations $x_{j, k}$ in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(i)}_j$. More generally we will use the notation $^{(\cdot)}$ to denote an ordering of values, e.g., denoting by $x^{\text{raw}, (i)}_j$ the $i$-th largest unsmoothed incidence value from season $j$.
\item[(c)] \textbf{Selection of reference set:} Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_j^{(i)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item[(d)] \textbf{Data transformation:} Apply a monotonically increasing transformation $y_j^{(i)} = f(x_j^{(i)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
\item[(e)] \textbf{Fitting a normal distribution:} Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates of its mean and standard deviation,
\begin{align}
\begin{split}
\bar{y} & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n y_j^{(i)}}{nm}\\
\sd & = \sqrt{ \frac{\sum_{j = 1}^m\sum_{i = 1}^n (y_j^{(i)}  - \bar{y})}{nm - 1}}.
\end{split}\label{eq:moments}
\end{align}
\item[(f)] \textbf{Computation of thresholds:} Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\bar{y}, \sd^2)$, i.e.\ compute
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd, \label{eq:def_q}
\end{equation}
where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution. The $\hat{q}_{Y, \alpha}$ can be interpreted as estimates of quantiles $q_{Y, \alpha}$ of an underlying theoretical distribution. A common choice for $\alpha$ is
\begin{itemize}
\item[(i)] the 40th percentile $\hat{q}_{Y, 0.4} = \bar{y} - 0.25 \sd$ as the threshold for medium intensity;
\item[(ii)] the 90th percentile $\hat{q}_{Y, 0.9} = \bar{y} + 1.28 \sd$ as the threshold for high intensity;
\item[(iii)] the 97.5th percentile $\hat{q}_{Y, 0.975} = \bar{y} + 1.96\sd$ as the threshold for very high intensity.
\end{itemize}
\item[(g)] \textbf{Transformation of thresholds to the original scale:} Obtain corresponding thresholds on the original scale by applying the inverse transformation, i.e. setting % $\hat{q}_{X, 0.4} = f^{-1}(\hat{q}_{Y, 0.4})$, $\hat{q}_{X, 0.9} = f^{-1}(\hat{q}_{Y, 0.9})$, $\hat{q}_{X, 0.975} = f^{-1}(\hat{q}_{Y, 0.975})$.
$$
\hat{q}_{X, \alpha} = f^{-1}(\hat{q}_{Y, \alpha})
$$
for $\alpha = 0.4, 0.9, 0.975$.
\item[(h)] \textbf{Application of thresholds:} The obtained thresholds are applied to classify the season peak value of the $(m + 1)-th$ season. Depending on the exact specification, thresholds can be applied either to the raw peak value $x_{m + 1}^{\text{raw}, (1)}$ or the smoothed version $x_{m + 1}^{\text{smo}, (1)}$.
\end{enumerate}

@@

The implementation of the WHO method in the publicly available \textit{WHO Average Curves} application \citep{WHO2023} likewise offers a lot of flexibility; our description is based on the default settings as of May 2023, as well as the description in \cite{WHO2014}. Smoothing of data prior to the computation of thresholds is recommended, with a default of $l = 3$ (adapted from a recommendation of $l = 4$ in \citealt[p68]{WHO2014}). Subsequently, $n = 1$ observation per season is used and by default no transformation is applied to the reference set. If peak incidences vary strongly across seasons, a log transformation is recommended. At least three historical seasons are required to compute thresholds, but it is noted that the ``accuracy of these thresholds should be expected to increase with the number of seasons of good quality data available'' \cite[p22]{WHO2023}. Thresholds are by default applied to unsmoothed new season peak values.

@@

To improve our understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search of articles published in English language and citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until May 2023 (identified via \textit{CrossRef} and \textit{Google Scholar}). The results are summarized in Table \ref{tab:literature}.

As can be seen from the large number of entries from the years 2019 and 2020, the MEM has quickly become a standard approach in the determination of intensity thresholds for influenza and other respiratory diseases. The articles come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies. In most analyses, threshold levels at the 40th, 90th and 97.5th percentile as in Section \ref{sec:definitions} are used. An interesting exception is the paper by \cite{Domegan2022} who introduced an additional \textit{super-extraordinary} threshold (though without specifying the corresponding quantile level). Variability with respect to the number $m$ of historical seasons included is considerable, with a range from 3 to 16 seasons. Consequently, the number $n$ of observations included per season ranges from two to ten. Only one of the reviewed papers, \cite{Dahlgren2022}, indicated a modification of the default setting $n = 30/m$  and used $n = 3$ with $m = 7$.

We only found six published applications of the WHO method, two of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020, Teeluck2021}. We note that in most applications of the WHO method it is not documented whether smoothing was applied and which value was chosen for $l$. In terms of quantile levels, several papers used the 95\% rather than 97.5\% quantile for the highest threshold.

@@

\caption{\footnotesize Published applications of the MEM and WHO methods to estimate intensity thresholds; $m$ is the number of historical seasons used, while $n$ is the number of observations used per season (in the bottom table we always have $n = 1$). The window width for smoothing with a moving average is denoted by $l$ (in the top table we always have $l = 1$). ``Percentiles'' indicates which percentiles were used for the medium, high and very high thresholds, with ``?'' indicating no information was found. Abbreviations: ARI/SARI = (severe) acute respiratory infection; ILI = influenza-like illness; RSV = respiratory syncytial virus.}

@@

% \vspace{-3mm}

@@

Maritius & ARI & 2012--2018 & varies$^\dagger$ & varies & 40, 90, 97.5 & Teeluck et al (\citeyear{Teeluck2021}) \\

@@

@@

\bottomrule
\multicolumn{7}{c}{$^\dagger$ Teeluck at al (\citeyear{Teeluck2021}) use various methods to split up multi-peak seasons, leading to different effective numbers of waves used.}\\ 

@@

We first consider the impact of the number $n$ of observations used per historical season. We assume that no smoothing is applied in Step (a) of the algorithm described in Section \ref{sec:definitions}. Denote by $\tilde{\mathbf{Y}}_j$ the random vector of the $n$ largest transformed incidence values from season $j$ in decreasing order, i.e.

@@

We assume that $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_m$ are identically and independently distributed, denoting their theoretical mean and covariance matrix by

@@

To make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. Construction of thresholds is based on the reference set $\mathcal{Y}$, which pools the elements of the vectors $\tilde{\mathbf{Y}}_1, \dots, \tilde{\mathbf{Y}}_m$. It can be shown that the expectations of the empirical mean $\bar{Y}$ and variance $S^2$ of the observations in the reference set (see in equation \eqref{eq:moments}) are

@@

where the second equality holds only if the transformed peak values $Y_{j}^{(1)}$ indeed come from a normal distribution.  This is a desirable property as our estimator $\hat{q}_{Y, \alpha}$ of $q_{Y, \alpha}$ is approximately unbiased. In expectation, thresholds on the transformed scale are thus such that the nominal threshold exceedance probabilities (60\%/10\%/2.5\%) are achieved. If a different value $n > 1$ is chosen, this will generally no longer be the case. Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $\hat{q}_{Y, \alpha}$ can then be expected to differ from $q_{Y, \alpha}$. By the definition of $\mu_i$ as the expectation of the $i$-th largest observation in a given season, $\mathbb{E}(\bar{Y})$ decreases in $n$. While the expected thresholds also depend on $\mathbf{\Sigma}$, this downward bias will usually translate to the $\hat{q}_{Y, \alpha}$. As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity. Intuitively speaking, including observations which are close to peaks, but not actually peaks, dilutes the reference set and pulls thresholds downward. When choosing $n = 30/m$ as suggested for the MEM, thresholds will tend to increase the more years of data are used to compute thresholds, and the probability of exceeding thresholds will decrease (approaching the nominal values from above).

@@

\subsection{Smoothing of time series prior to computing thresholds}
\label{subsec:smoothing}

Next we assess the role of smoothing historical data prior to computing thresholds. As derivations get tedious otherwise, we only consider the case where $f$ is the identity function, i.e., no transformation is applied to the reference set and thus $
Y^{(i)}_j = X^{(i)}_j$. Also, as smoothing and multiple observations per season are generally not used in parallel we assume $n = 1$. Now denote by $p_j$ the peak week of the smoothed incidence in season $j$ such that
\begin{equation}
Y_j^{(1)} =  X_j^{(1)} = X^{\text{smo}}_{j, p_j} = \frac{1}{l} \sum_{d = 0}^{l - 1} X^{\text{raw}}_{j, p_j - d}.\label{eq:Y_1_smoothing}
\end{equation}
Moreover, we define a random vector containing the observations $X^{\text{raw}}_{j, p_j}, \dots, X^{\text{raw}}_{j, p_j - l + 1}$ which average to $Y_j^{(1)}$ in decreasing order,
$$
\tilde{\mathbf{X}}^\text{raw}_j = \begin{pmatrix} \tilde{X}_j^{\text{raw}, (1)} & = \ \ \max(X^{\text{raw}}_{j, p_j}, \dots, X^{\text{raw}}_{j, p_j - l + 1}) \\ \vdots & \vdots \\ X_j^{\text{raw}, (l)} & = \ \ \min(X^{\text{raw}}_{j, p_j}, \dots, X^{\text{raw}}_{j, p_j - l + 1}) \end{pmatrix}.
$$
The theoretical mean and covariance matrix of $\tilde{\mathbf{X}}^\text{raw}_j$ are denoted by
\begin{align}
\mathbb{E}\left(\tilde{\mathbf{X}}^\text{raw}_j\right) = \boldsymbol{\mu}^\text{raw} = \left(\begin{array}{c}
\mu^\text{raw}_1\\
\vdots\\
\mu^\text{raw}_l
\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}\left(\tilde{\mathbf{X}}^\text{raw}_j\right) = \boldsymbol{\Sigma}^\text{raw} =
\left(\begin{array}{ccc}
\sigma^\text{raw}_{1, 1} & \cdots & \sigma^\text{raw}_{1, l}\\
\vdots & \ddots &\vdots\\
\sigma^\text{raw}_{l, 1} & \cdots & \sigma^\text{raw}_{l, l}
\end{array}\right),
\end{align}
respectively. Note that by construction we have $\mu^\text{raw}_1 \geq \mu^\text{raw}_2 \geq \dots > \mu^\text{raw}_l$. As we assumed $n = 1$, the reference set $\mathcal{Y}$ consists simply of $Y^{(1)}_1, \dots, Y^{(1)}_m$ as given in equation \eqref{eq:Y_1_smoothing}. It is straightforward to show that in this case the expectations of $\bar{Y}$ and $S^2$ are
\begin{align}
\mathbb{E}(\bar{Y}) = \frac{1}{l} \sum_{i = 1}^l \mu^\text{raw}_i, \ \ \ 
\mathbb{E}(S^2) = \frac{1}{l^2} \sum_{i = 1}^l \sum_{i' = 1}^l \sigma^\text{raw}_{i', i}.
\label{eq:expectation_moments_smoothing}
\end{align}
These can be plugged into equation \eqref{eq:expectation_q} to approximate the expected thresholds. Similarly to the previous section, for $l = 1$ we obtain
\begin{equation}
\mathbb{E}(\hat{q}_{Y, \alpha}) \approx \mathbb{E}(\bar{Y}) + z_\alpha \sqrt{\mathbb{E}(S^2)} \approx \mu^\text{raw}_1 + z_\alpha \sigma^\text{raw}_1.
\label{eq:expected_threshold_smooth}
\end{equation}
If in addition the normality assumption holds and $q_{Y, \alpha} = \mu_1^\text{raw} + z_\alpha \sigma_1^\text{raw}$, then equation \eqref{eq:expected_threshold_smooth} implies that the threshold $\hat{q}_\alpha$ is again approximately unbiased. % in the sense that in expectation they are such that the unsmoothed peak of a new season will exceed them with the intended probabilities (60\%/10\%/2.5\%).
For $l > 1$ we get by construction $\mathbb{E}(\bar{Y}) \leq \mu^\text{raw}_1$ and usually also $\mathbb{E}(\sqrt{S^2}) \leq \sigma^\text{raw}_1$. In practice both are likely to hold in strict inequality. The high and very high thresholds will thus be lower than for $l = 1$ and exceeded more frequently by unsmoothed new peaks. If thresholds are applied to new smoothed peak values, though, this problem is avoided as the mean and variance of the smoothed peak values correspond to the expressions from equation \eqref{eq:expectation_moments_smoothing}.

As in Section \ref{subsec:choice_n}, the key to ensuring calibrated thresholds is that the reference set need to be comparable to new peaks; both smoothed peak values and values which are close to, but not actually peaks are systematically different from new unsmoothed peak values. They are thus not suitable as reference values, and the resulting thresholds will be biased. While we did not adapt our derivation to the case where data are log-transformed prior to computing thresholds, our simulation results in Section \ref{subsec:sim_smoothing} confirm that the described qualitative patterns likewise hold.
